The evolution of gene expression in the term placenta of viviparous mammals

### 2015-11-03 manuscript/manuscript.Rnw
discuss go enrichement more

\subsection{DAVID Analyses of Individual Transcriptomes}

In order to identify major functional categories of genes expressed in
the individual transcriptomes, we undertook functional annotation
clustering analyses using DAVID (see Methods). Concentrating on genes
with relatively high expression values, we conducted DAVID analyses
using a list of genes with \ac{FPKM} values included within the top 5\%
distribution of expressed genes. Functional annotation clusters from
all six placenta transcriptomes show enrichment in similar clusters of
genes (Supplemental Table X). Significant enrichment in genes (ES>1.3,
p>0.05) in the top 5\% of \ac{FPKM} values in at least five transcriptomes
included clusters such as ribosome, vesicle, cytoskeleton,
glycoprotein, lysosome, annexins, hormone stimulus or response,
growth, calcium binding, cell migration, basement membrane,
antioxidant activity or oxidative stress, cell adhesion, and
mitochondrion among others (Supplemental Table X).

We also investigated enrichment in functional annotation clusters in
the top 100 genes of each transcriptome (Table 9). As in the analyses
of the top 5\% of genes, the top 100 genes show significant enrichment
(ES>1.3) in all six transcriptomes in genes associated with ribosomes.
Human shows enrichment in clusters such as collagen-binding, hormone
activity, female pregnancy, coagulation, and the defense response.
Bonobo shows enrichment in clusters related to apoptosis,
cytoskeleton-related genes, and the stress response. Cow shows only
four significantly enriched clusters; however many genes highly
expressed in the cow are ruminant-specific placenta genes and are not
yet annotated to specific GO categories. Dog shows enrichment in genes
associated with RNA, as well as homeostasis, glycoproteins, lyosome,
and calcium binding. Elephant shows enrichment in genes associated
with RNA, coagulation and calcium binding. Opossum shows enrichment in
genes involved in apoptosis, oxidative stress, mitochondrion, hormone
metabolism, and lipids.

\begin{table}
<<go.analysis.all.tissue,echo=FALSE,results="asis">>=
go.analysis.wide <- 
    data.table(dcast(data=go.results,
                     formula=GO.ID+type+Term+Ontology~species,
                     value.var="FDR"))
go.analysis.wide <- 
    go.analysis.wide[,`max(FDR)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],
                                    1,max,na.rm=TRUE)]
go.analysis.wide.table <- 
    go.analysis.wide[`max(FDR)` <= 0.1 & grepl("No Housekeeping",type),
                     list(GO.ID,Term,Ontology,`max(FDR)`)]
go.analysis.wide.table <- 
    go.analysis.wide.table[order(`max(FDR)`)]
print(xtable(go.analysis.wide.table,
             display=c("d","s","s","s","e")
             ),
      booktabs=TRUE,
      floating=FALSE
      )
@ 
\caption{GO terms with enrichment in all species analyzed with
  housekeeping genes removed}
\label{tab:go_terms_all_species}
\end{table}

>>>>

\subsection{topGO Analyses of Individual Transcriptomes}

In order to identify major functional categories of genes expressed in
the individual transcriptomes, we undertook functional annotation
clustering analyses using topGO. We examined \ac{GSE} in 1:1 human
orthologs which had \ac{FPKM} values in the top 5\% of expressed genes
%
Significant enrichment in genes ($\mathrm{FDR}\le0.05$) in the top 5\%
of \ac{FPKM} values in at least 12 of the
\Sexpr{length(species.names)} species included clusters such as
membrane organization , locomotion, tissue development, extracellular
matrix organization, regulation of peptidase activity, and various
cell death related pathways among others
(\cref{tab:go_terms_all_species}).

(Supplemental %
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}.) %

Human shows enrichment in clusters such as collagen-binding, hormone
activity, female pregnancy, coagulation, and the defense response.
Bonobo shows enrichment in clusters related to apoptosis,
cytoskeleton-related genes, and the stress response. Cow shows only
four significantly enriched clusters; however many genes highly
expressed in the cow are ruminant-specific placenta genes and are not
yet annotated to specific GO categories. Dog shows enrichment in genes
associated with RNA, as well as homeostasis, glycoproteins, lyosome,
and calcium binding. Elephant shows enrichment in genes associated
with RNA, coagulation and calcium binding. Opossum shows enrichment in
genes involved in apoptosis, oxidative stress, mitochondrion, hormone
metabolism, and lipids.

\begin{table}
  \footnotesize
<<go.analysis.all.tissue,echo=FALSE,results="asis">>=
go.analysis.wide <- 
    data.table(dcast(data=go.results,
                     formula=GO.ID+type+Term+Ontology~species,
                     value.var="FDR"))
go.analysis.wide <- 
    go.analysis.wide[,`n(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
                                         function(x){sum(x <= 0.05)})]
go.analysis.wide <- 
    go.analysis.wide[,`max(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
                                           function(x){if (!any(is.finite(x) & x<=0.05)) {
                                                           return(NA)
                                                       } else {
                                                           max(x[is.finite(x) & x<=0.05],na.rm=TRUE)
                                                       }})]
go.analysis.wide.table <- 
    go.analysis.wide[!is.na(`n(FDR ≤ 0.05)`) & `n(FDR ≤ 0.05)` >= 12 &
                         grepl("Top 5% No Housekeeping",type),
                     list(GO.ID,Term,Ontology,`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
go.analysis.wide.table <- 
    go.analysis.wide.table[order(Ontology,-`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
setnames(go.analysis.wide.table,"GO.ID","GO ID")

print(xtable(go.analysis.wide.table,
             display=c("d","s","s","s","d","e"),
             align=  c("c","c","c","c","c","c")
             ),
      math.style.exponents=TRUE,
      include.rownames=FALSE,
      booktabs=TRUE,
      floating=FALSE
      )

@ 
\caption{GO terms with significant enrichment ($\mathrm{FDR}\le0.05$)
  in 12 or more of the species analyzed from genes in the top 5\% of
  expression. $n(\mathrm{FDR}\le0.05)$ is the number of species with
  significant enrichment ($\mathrm{FDR}\le 0.05$) in the GO term
  listed. $max(\mathrm{FDR}\le0.05)$ is the maximum FDR value in all
  species with an FDR of less than $0.05$.}
\label{tab:go_terms_all_species}
\end{table}
